Finiteness, Cardinality, and Combinatorics in Homotopy Type Theory

### 2020-01-04 paper.tex
finally compiles

\input{preamble}
\title{Finiteness, Cardinality, and Combinatorics in Homotopy Type Theory}
\author{Donnacha Oisín Kidney}
\begin{document}
\maketitle
\listoftodos
\tableofcontents
\newcommand{\pifun}[2]{\prod_{\left( \mathit{ #1 } : #2 \right)}}
\begin{quote}
  \textbf{Abstract}
  We explore five notions of finiteness in Homotopy Type Theory \cite{hottbook}.
  We prove closure properties about all of these notions, culminating in a proof
  that decidable Kuratowski-finite sets form a topos.

@@

  We formalise our work in Cubical Agda
  \cite{vezzosiCubicalAgdaDependently2019}, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{quote}
\section{Introduction}
Proofs in a constructive setting are more substantial things than their
classical counterparts.
Rather than just evidence for some fact, they are objects in their own right,
which we can manipulate and extract information from.

@@

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsubsection{Equivalence}
While manifest Bishop finiteness may seem stricter than split enumerability, we
will see soon that it is actually equivalent in strength.
Why study it separately to split enumerability at all, then?
What purpose does it serve on our road to Kuratowski finiteness?
Those questions are answered by the following theorem:
\begin{lemma} \label{bishop-equiv}
  Any manifestly Bishop finite type has an equivalence with a finite prefix of
  the natural numbers.
\end{lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite[Definition~4.4.1]{hottbook}. \todo[noline]{Need this in appendix also}
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.

@@

Finally, we can prove isomorphisms between the curried and uncurried versions of
functions:
\ExecuteMetaData[agda/Data/Product/N-ary.tex]{curry-n-iso}
And dependent functions:
\ExecuteMetaData[agda/Data/Product/N-ary.tex]{curry-n-pi-iso}
With these combinators, we can implement the search functions in an
arity-generic way:
\ExecuteMetaData[agda/Cardinality/Finite/Search.tex]{all-nary}
And automate away our proofs:
\ExecuteMetaData[agda/Data/Pauli.tex]{prf-curried}

\begin{subappendices}

\section{Notation, Cubical Type Theory, and Agda}
Our chosen setting is Cubical Type Theory (CuTT)
\cite{cohenCubicalTypeTheory2016}, which is closely related to Homotopy Type
Theory (HoTT) \cite{hottbook}.
We will go through the basics and some notation here.
\subsection{Martin-Löf Type Theory}
CuTT has its roots in Martin-Löf intuitionistic type theory
\cite{martin-lofIntuitionisticTypeTheory1980}.
We will describe our notation for the MLTT parts of CuTT here (although our
notation is quite standard).

The empty or \(\mathbf{0}\) type will be denoted by \(\bot\), and the
\(\mathb{1}\) type by \(\top\).
The single inhabitant of \(\top\) is called \(\text{tt}\).

``not'' is expressed as follows:
\begin{equation}
  \neg A = A \rightarrow \bot
\end{equation}
In other words, to say a type is impossible we say we can derive \(\bot\) from
it.

We also use the \(\mathbf{2}\) type, which we call \(\textbf{Bool}\).
Its two canonical inhabitants are \(\text{true}\) and \(\text{false}\).

We will use the term type family (or simply ``family'') to mean a non-dependent
function with a codomain in Type.

Dependent function types, or \(\prod\)-types, use the following notation:
\begin{equation}
  \prod_{(x : A)} , B (x)
\end{equation}
In Agda we use a single arrow to denote the same:
\ExecuteMetaData[agda/Snippets/Introduction.tex]{pi}
When the type of the left-hand-side can be inferred, we are permitted to use the
following notation:
\ExecuteMetaData[agda/Snippets/Introduction.tex]{forall}
For non-dependent functions we use the single arrow (\(\rightarrow\)).

Dependent pairs, or \(\sum\)-types, use a similar notation to \(\prod\)-types.
\begin{equation}
  \sum_{(x : A)} , B (x)
\end{equation}
In Agda:
\ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma}
Again, we get a shorthand when the type of the first element can be inferred:
\ExecuteMetaData[agda/Snippets/Introduction.tex]{exists}
For non-dependent pairs we use the \(\times\) symbol.
We access the two fields of \(\sum\) types with \(\text{fst}\) and
\(\text{snd}\):
\begin{align}
  \text{fst} &: \sum_{(x : A)} , B (x) \rightarrow A \\
  \text{snd} &: \prod_{(x : \sum_{(y : A)} , B (y))} , B (\text{fst}(x))
\end{align}

Finally, the identity type is denoted by the triple equals (\(\equiv\)).
The normal equals (\(=\)) is reserved for definitions.

\subsection{Agda Basics}
Agda \cite{norellDependentlyTypedProgramming2008} is a dependently-typed
functional programming language, with roots in Haskell and ML.
The core type theory is based on MLTT, though support for CuTT was added in the
last year \cite{vezzosiCubicalAgdaDependently2019}.
\subsubsection{Declarations}
The anatomy of a simple Agda declaration is thus:
\ExecuteMetaData[agda/Snippets/Introduction.tex]{abool}
We first have a type signature: the name of the variable is to the left of the
colon, and its type to the right.
Below, we give an equation: as Agda is pure, it suffices to imagine that
everywhere the name on the left of the equals sign is seen, we can replace it
with the value on the right.
\subsubsection{Functions}
Functions are usually declared with pattern matching.
\ExecuteMetaData[agda/Snippets/Introduction.tex]{not}
Again, each clause is an equation.
Agda's purity will ensure that these clauses hold true definitionally\footnotemark.
\footnotetext{
  Actually, it is possible to construct functions whose equations
  hold propositionally, but not definitionally.
  An example might be the following:
  \ExecuteMetaData[agda/Snippets/Introduction.tex]{non-exact-split}
  This is actually attempted by a lot of Agda beginners: often the compiler will
  complain ``couldn't match type \(n\) with \(n + 0\)'', so the programmer will
  simply add it as a clause.
  Unfortunately, this actually over complicates things in the long run.
  Neither the second nor third clause here will be definitionally equal: if you
  have an obligation \(n \oplus 0 \qeq n\), the first pattern (\(0 \oplus m = m\))
  has to be ruled out first, which means you'll have to do a manual pattern
  match in your proof.
  Similarly for the third pattern: to prove the obligation (\((1 + n) \oplus m
  \qeq 1 + (n \oplus m)\)), \emph{both} preceding patterns have to be checked
  against first.

  This ordering requirement is introduced by overlapping patterns, and Agda
  warns against it.
  Knowing when to avoid it and when it doesn't matter is one of the aspects of
  programming and proving in Agda which is more of an art form.
}

As Agda is a functional language, functions can also be defined with lambda
expressions.
\ExecuteMetaData[agda/Snippets/Introduction.tex]{id}

Finally, ``mixfix'' operators play a large role in Agda's syntax.
We can have binary operators:
\ExecuteMetaData[agda/Snippets/Introduction.tex]{add}
Unary operators (here, a prefix increment operator):
\ExecuteMetaData[agda/Snippets/Introduction.tex]{unary-inc}
Ternary operators:
\ExecuteMetaData[agda/Snippets/Introduction.tex]{ifte}
And so on.
\subsubsection{Data Type Definitions}
We can define new data types in Agda by listing their constructors:
\ExecuteMetaData[agda/Snippets/Introduction.tex]{maybe}
\subsubsection{Termination}
\todo[inline]{Coinductive types, termination checking, etc.}
\subsubsection{Universe Levels}
All dependently-typed programming languages have to contend, at some point or
another, with Girard's paradox
\cite{girardInterpretationFonctionelleElimination1972}.
The paradox concerns the type of types: it's the type-theory analogue of
Russell's paradox.

Unfortunately, most of the ways to avoid the paradox add significant clutter to
the type theory, and increase the complexity of even simple declarations.
Agda's answer is no exception.

In Agda, every type is tagged with a ``universe level''.
The basic types like \(\AgdaDatatype{Bool}\) live in universe level 0, the type
of types lives in universe level 1, the type of types in universe level 1 lives
in universe level 2, and so on.
We won't go into it much more than that: usually there is an obvious way to
handle the universe level of a given expression ``properly'', and cases where
universe levels do something interesting come up rarely (although
section~\ref{practical} does touch on one).
We only mention it here to explain the expressions like \(A :
\AgdaDatatype{Type} \; a\) which show up in many examples: \(a\) there is the
universe level of the type \(A\).
\subsection{Cubical Basics}
This will not serve as an introduction CuTT,
but rather a brief overview of the parts of the theory we will be using.
For a fuller reference we direct the reader first to the ``HoTT Book''
\cite{hottbook}, then the paper on Cubical Type Theory
\cite{cohenCubicalTypeTheory2016}, and finally the paper on Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
\subsubsection{Paths}
Central to CuTT is the \emph{path} type.
This is a path in the topological sense, although we use it as a representation
for equality (which makes sense: central to HoTT is the
equivalence of these two concepts).
In most of the code presented here this will not behave in any different way to
the traditional identity type in Martin-Löf type theory: one exception to this
is pattern-matching, which is not available on paths in Cubical Agda currently.
As a result, we have to use its eliminator directly (if needed), although
usually higher-level combinators like \(\AgdaFunction{subst}\) are sufficient.

We will use \(\equiv\) as a notation for path types as well as the identity
types.
Although an identity type does exist in Cubical Type Theory, we will not use it,
so we do not fear much confusion on this point.
\subsubsection{Homotopy Levels}
Next, we have the concept of ``homotopy levels''.
These classify types in homotopy type theory according to their degree of
truncation.
Homotopy level 1, for instance, is the ``mere propositions'': the types for
which every element is equal to every other.
As a predicate, we can define \(\AgdaFunction{isProp}\) like so:
\begin{equation}
  \text{isProp}(A) = \prod_{(x : A)} , \prod_{(y : A)} , x \equiv y
\end{equation}
In Agda:
\ExecuteMetaData[agda/Snippets/Introduction.tex]{isProp}

One level up from the propositions are the (homotopy) sets.
These are types for which all paths are themselves propositions.
\begin{equation}
  \text{isSet}(A) = \prod_{(x : A)} , \prod_{(y : A)} , \text{isProp}(x \equiv y)
\end{equation}
In Agda:
\ExecuteMetaData[agda/Snippets/Introduction.tex]{isSet}

Finally, one level down from the propositions are the contractions.
These are types for which every element is equal to some specific element.
\begin{equation}
  \text{isContr}(A) = \sum_{(x : A)} , \prod_{(y : A)} , x \equiv y
\end{equation}
\ExecuteMetaData[agda/Snippets/Introduction.tex]{isContr}
\subsubsection{Univalence}
\todo[inline]{Flesh out}
To describe this we first need fibers:
\begin{equation}
  \text{fiber}_f(y) = \sum_{(x : A)} , f (x) \equiv y
\end{equation}
\ExecuteMetaData[agda/Snippets/Introduction.tex]{fiber}
From this we build the notion of an ``equivalence'': a function whose fibers are
all contractions.
\begin{equation}
  A \simeq B = \sum_{(f : A \rightarrow B)}, \prod_{(y : B)} , \text{isContr}(\text{fiber}_f(y))
\end{equation}
\ExecuteMetaData[agda/Snippets/Introduction.tex]{equiv}
The univalence axiom is that all equivalences are equivalent to the identity
type.
\begin{definition}[The Univalence Axiom]
  \begin{equation}
    \simeq \; \simeq \; \equiv
  \end{equation}
\end{definition}
In practice this means that we can prove equivalence between two types and then
transport code on one to code on the other.
It also means that isomorphisms can be upgraded to fully-fledged equalities.
\begin{definition}[Isomorphism]
  An isomorphism between two types \(A\) and \(B\) is a pair of functions:
  \begin{align}
    f : A \rightarrow B \\
    g : B \rightarrow A
  \end{align}
  Such that they are inverses of each other.
  \begin{align}
    \prod_{(x : A)} , g(f(x)) \equiv x \\
    \prod_{(y : B)} , f(g(y)) \equiv y
  \end{align}
\end{definition}
\ExecuteMetaData[agda/Snippets/Introduction.tex]{iso}
\subsubsection{Higher Inductive Types}
Along with univalence, the other significant benefit we get from path types is
the ability to define new paths on types.
A direct use of this feature is synthetic topology.
We can, for instance, define the circle \(S^1\) as a higher-inductive type.
\ExecuteMetaData[agda/Snippets/Introduction.tex]{circle}

Another important use of higher inductive types is quotient types.
These types are unavailable in ML type theory, but in Cubical Type Theory we can
declare a quotient (here, the type of Kuratowski finite sets) like so:
\ExecuteMetaData[agda/Algebra/Construct/Free/Semilattice.tex]{def}
The first two constructors are standard data constructors, which mimic those for
the list in Haskell.
The next two add the equations we want our type to obey: we don't want it to
care about order or duplicates.
The last constructor \emph{truncates} the type, ensuring that it is a set.
If we hadn't included it paths would not be mere propositions: there would be
more than one path between any given elements of a type.
The ability to have multiple paths is what allows Cubical Type Theory to do the
more advanced things in topology etc., but we don't need its power here.

One HIT in particular will be of great use to us: the HIT for propositional
truncation, denoted with \(\lVert \cdot \rVert\).
We will go through it in detail in section~\ref{prop-trunc}.
\section{Supplementary Definitions}
This section will serve as a reference for any definitions used in the text
which are either not important enough to be examined inline, or common enough to
justify elision.
Many of the definitions come straight from the ``HoTT Book'' \cite{hottbook}.
\subsection{Decidable Types}
It is important to be able to talk about ``decisions'' for things like decidable
equality, and so on.
\begin{definition}[Decision]
  A decision for some type \(P\) is represented by the type \(\mathbf{Dec}(A)\),
  which has two constructors:
  \begin{equation}
    \begin{alignedat}{2}
      \text{yes} &:      &A \rightarrow \mathbf{Dec}(A) \\
      \text{no}  &: \neg &A \rightarrow \mathbf{Dec}(A)
    \end{alignedat}
  \end{equation}
\end{definition}

Internally, we actually use a rather more complicated type, which has preferable
computational behaviour.
The interface is the same, however.
\subsection{Propositional Truncation} \label{prop-trunc}
Essentially, a value of type \(\lVert A \rVert\) proof that an \(A\) exists,
without revealing which \(A\) in particular.
In HoTT, it allows us to work in a setting similar to classical logic.
In Agda its definition is:
\ExecuteMetaData[agda/Snippets/Introduction.tex]{prop-trunc}
Any eliminators from this type must not distinguish between different values of
\(A\).

\begin{lemma}[Propositional Elimination]
  We can eliminate from the type \(\lVert A \rVert\) to \(B\) with a function
  \(A \rightarrow B\) if B is a proposition.
\end{lemma}
\begin{proof}
  By case-analysis on \(\lVert A \rVert\).
  In the first case, we apply our function.
  In the second, we are required to show that that application agrees on any two
  values of \(\lVert A \rVert\) \(\lVert A \rVert\).
  This is true as \(B\) is a proposition.
  In Agda:
  \ExecuteMetaData[agda/HITs/PropositionalTruncation/Proofs.tex]{rec}
\end{proof}

Since \(\lVert A \rVert\) is itself a proposition, we can work ``under'' the
truncation as much as we like.
In fact the type \(\lVert \cdot \rVert\) itself forms a monad, allowing us to
use monadic notation.
\begin{lemma}
  \(\lVert \cdot \rVert\) is a monad, with bind being the eliminator defined
  above, and return being the \(\lvert \cdot \rvert\) constructor.
\end{lemma}
\begin{proof}
  The monad laws are satisfied trivially, as \(\lVert \cdot \rVert\) is a
  proposition.
\end{proof}

We can further eliminate into a type that isn't a proposition, as long as we
prove that the function used in the eliminator doesn't discriminate between
different \(A\)s in \(\lVert A \rVert\).
\begin{lemma}[Propositional Elimination by Coherence] \label{elim-non-prop}
  We can eliminate from a propositionally-truncated type into a non-proposition,
  as long as the function is coherently constant
  \cite{krausGeneralUniversalProperty2015}.
\end{lemma}
\begin{proof}
  \todo[inline]{sketch proof}
\end{proof}

From these eliminators, we can derive some useful lemmas.
\begin{lemma}[Propositional Refutation]
  While a propositionally truncated proof is not enough to prove the
  non-truncated type, it is enough to \emph{refute} the non-truncated negation
  of the type.
  \begin{equation}
    \lVert A \rVert \rightarrow \neg (\neg A)
  \end{equation}
\end{lemma}
\begin{proof}
  We will use propositional elimination.
  First, to go from \(A\) to \((A \rightarrow \bot) \rightarrow \bot\) is
  trivial.
  Secondly, to show that \((A \rightarrow \bot) \rightarrow \bot\) is a
  proposition, we can use the fact that \(\bot\) is a proposition, and then
  derive (via function extensionality) that any \(A \rightarrow \bot\) is also a
  proposition.
\end{proof}

\begin{lemma}[Recomputation]
  If a type \(A\) is decidable, and there exists a proof of the type \(\lVert A
  \rVert\), we can derive a proof of \(A\).
  In other words, the following function exists:
  \begin{equation}
    \text{recompute} : \textbf{Dec}(A) \times \lVert A \rVert \rightarrow A
  \end{equation}
\end{lemma}
\begin{proof}
  By case analysis of the decision over \(A\), we know that the false case must
  be untrue (since we can refute it using propositional refutation), and we can
  therefore extract the true case.
\end{proof}

Occasionally we will call families ``predicates'' when we mean to treat the
family as a property on the domain.
Formally this might imply that the codomain is a proposition (i.e. isomorphic to
either \(\top\) or \(\bot\)), but it is not always necessary to prove this fact,
and indeed occasionally we will want to use some of the information stored in
the predicate, making it not propositional.
For instance, a natural number being ``not prime'' we would refer to as a
predicate, but the proof of this fact may be a list of the number's prime
factors, so we will not truncate it away.

When we say that two predicates \(P\) and \(Q\) are ``equal strength'', we mean
there are a pair of functions \(P \rightarrow Q\) and \(Q \rightarrow P\).
This is distinct from \(P\) and \(Q\) being isomorphic, as we haven't shown
anything regarding the relationship between those two functions.
However, if \(P\) and \(Q\) are themselves propositions, this does indeed imply
an isomorphism.
\begin{lemma} \label{equal-strength}
  For any two types \(A\) and \(B\), if there are functions \(f : A \rightarrow
  B\) and \(g : B \rightarrow A\), \(\lVert A \rVert\) and \(\lVert B \rVert\)
  are equivalent.
\end{lemma}
\begin{proof}
  We prove equivalence by way of isomorphism.

  Let \(A\), \(B\), \(f\), and \(g\) be defined as above.
  For isomorphism, we first provide a pair of functions \(\lVert f \rVert\) and
  \(\lVert g \rVert\), which are defined by lifting \(f\) and \(g\) under
  propositional truncation, respectively.

  Next, we need to prove that \(g\) is a section of \(f\).
  \begin{equation}
    \text{section}(f, g) = f \circ g \equiv id
  \end{equation}
  Since \(\lVert B \rVert\) is a proposition, this is true by function
  extensionality.

  Finally, \(f\) is a retraction of \(g\).
  \begin{equation}
    \text{retraction}(f, g) = g \circ f \equiv id
  \end{equation}
  Again, since \(\lVert A \rVert\) is a proposition, this is true by function
  extensionality.

  In Agda:
  \ExecuteMetaData[agda/HLevels/Proposition.tex]{prop-iff}
\end{proof}
\subsection{Functions}
\subsubsection{Surjections}
The following is the definition of surjectivity in HoTT:
\begin{equation}
  \text{Surjective}(f) = \prod_{(y : B)} , \lVert \text{fiber}_f(y) \rVert
\end{equation}
\ExecuteMetaData[agda/Function/Surjective.tex]{surj-def}
We are also interested in \emph{split} surjectivity, where we don't truncate the
fiber:

\begin{equation}
  \text{SplitSurjective}(f) = \prod_{(y : B)} , \text{fiber}_f(y)
\end{equation}
\ExecuteMetaData[agda/Function/Surjective.tex]{split-surj-def}

A surject\emph{ion}, then, is a function which is surject\emph{ive}:
\begin{equation}
  A \twoheadrightarrow B = \sum_{(f : A \rightarrow B)} \text{SplitSurjective}(f)
\end{equation}
\ExecuteMetaData[agda/Function/Surjective.tex]{surjection}
\subsubsection{Injections}
The standard definition of an injective function:
\ExecuteMetaData[agda/Function/Injective.tex]{injective-def}
An injection:
\ExecuteMetaData[agda/Function/Injective.tex]{injection-def}
\subsection{Fin}
\(\textbf{Fin}\) is a finite prefix of the natural numbers.
Semi-formally:
\begin{equation}
  \textbf{Fin}(n) = \left[ 0 \dots n \right)
\end{equation}
There are actually a number of ways to define \(\textbf{Fin}\) in type theory.
The method we use here is as follows:
\ExecuteMetaData[agda/Data/Fin.tex]{def}
\subsection{Split Enumerability}
Here is how we define it in Agda:
\ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{definition}
\subsection{Predicates Over Lists}
These two well-known dual types for quantifying a predicate over a list are used
extensively in this work.
We will define them using the indexing operator \(!\), which has the following
type:
\begin{equation}
  ! : \prod_{(\mathit{xs} : \textbf{List}(A))} , \textbf{Fin}(\text{length}(\mathit{xs})) \rightarrow A
\end{equation}
For some predicate \(P\), the two definitions are as follows:
\begin{definition}[Any]
  \begin{equation}
    \Diamond P \mathit{xs} = \sum_{(i : \textbf{Fin}(\text{length}(\mathit{xs})))} , P(xs \ind i)
  \end{equation}
  \ExecuteMetaData[agda/Data/List/Relation/Unary.tex]{any-def}
\end{definition}
\begin{definition}[All]
  \begin{equation}
    \Box P \mathit{xs} = \prod_{(i : \textbf{Fin}(\text{length}(\mathit{xs})))} , P(xs \ind i)
  \end{equation}
  \ExecuteMetaData[agda/Data/List/Relation/Unary.tex]{all-def}
\end{definition}
We are using the tradition modal logic operators here: \(\Diamond\)
means ``any'', and \(\Box\) means ``all''.
\subsection{Membership}
Several other constructions are defined in terms of \(\AgdaDatatype{◇}\),
starting with \(\AgdaDatatype{∈}\):
\begin{equation}
  x \in \mathit{xs} = \Diamond (\equiv x) \mathit{xs}
\end{equation}
\ExecuteMetaData[agda/Data/List/Membership.tex]{in-def}

To show \emph{unique} membership, the membership proof itself must be a
contraction.
\begin{equation}
  x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
\end{equation}
\ExecuteMetaData[agda/Data/List/Membership.tex]{in-uniq}
\section{Supplementary Code}
\subsection{Permutations}
Swap:
\ExecuteMetaData[agda/Data/List/Relation/Binary/Permutation.tex]{swap}
Cons:
\ExecuteMetaData[agda/Data/List/Relation/Binary/Permutation.tex]{cons}
Tail:
\ExecuteMetaData[agda/Data/List/Relation/Binary/Permutation.tex]{tail}
\subsection{Non-Dependent Sum (Disjoint Union)}
\ExecuteMetaData[agda/Snippets/Introduction.tex]{sum-def}
\subsection{Split Enumerability}
Closure over \(\bot\):
\ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{empty}
Closure over \(\top\):
\ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{unit}
Closure over non-dependent sum:
\ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sum-prf}
\subsection{Lists}
\subsubsection{Indexing}
\ExecuteMetaData[agda/Data/List.tex]{indexing}
\subsubsection{Removing Duplicates}
\ExecuteMetaData[agda/Data/List/Membership.tex]{uniques}
\subsection{Functions}
If there is an injection from \(A\) into a set with decidable equality, \(A\)
has decidable equality.
\ExecuteMetaData[agda/Function/Injective.tex]{discrete}
If there is a surjection from \(A\) to \(B\), there is an injection from \(B\)
to \(A\).
\ExecuteMetaData[agda/Function/Surjective.tex]{surj-to-inj}
\subsection{Membership Proofs}
All iso:
\ExecuteMetaData[agda/Data/List/Relation/Unary.tex]{all-iso}
Any iso:
\ExecuteMetaData[agda/Data/List/Relation/Unary.tex]{any-iso}
A small function to make membership proofs shorter:
\ExecuteMetaData[agda/Data/List/Membership.tex]{at}
Pushing and pulling: going between \(x \; \AgdaDatatype{∈} \; xs\) and \(x
\AgdaDatatype{∈} \; y \; \AgdaInductiveConstructor{∷} \; xs\).
\ExecuteMetaData[agda/Data/List/Relation/Unary.tex]{push-pull}
Push-pull for unique proofs of membership.
\ExecuteMetaData[agda/Data/List/Relation/Unary.tex]{push-pull-unique}
Proving that an element at the head of a list is in that list.
\ExecuteMetaData[agda/Data/List/Relation/Unary.tex]{here-unique}
\(\AgdaFunction{map}\) distributes over \(\AgdaDatatype{∈}\)
\ExecuteMetaData[agda/Data/List/Relation/Unary.tex]{cong-any}
\ExecuteMetaData[agda/Data/List/Membership.tex]{cong-in}
An element is not in a list it's been removed from.
\ExecuteMetaData[agda/Data/List/Membership.tex]{not-in-removed}
\ExecuteMetaData[agda/Data/List/Membership.tex]{cons-remove}
If \(x\) isn't in a list, it is not in that list with another element removed.
\ExecuteMetaData[agda/Data/List/Membership.tex]{remove-other}
Removing another element from a list doesn't affect \(x\)'s proof of unique
membership.
\ExecuteMetaData[agda/Data/List/Membership.tex]{unique-remove-other}
If \(x\) is in \(xs\), \(x\) is uniquely in the unique elements of \(xs\).
\ExecuteMetaData[agda/Data/List/Membership.tex]{unique-preserves}
% \subsection{Convolution}
% Convolution of lists.
% \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{convolve}
% Convolution of streams.
% \ExecuteMetaData[agda/Cardinality/Infinite.tex]{convolve}
% Convolution of colists.
% \ExecuteMetaData[agda/Cardinality/PossiblyInfinite.tex]{convolve}
% \subsection{Decidable Predicates}
% Rather than the ``usual'' formulation of \(\AgdaDatatype{Dec}\), we use the
% following definition:
% \ExecuteMetaData[agda/Relation/Nullary.tex]{dec-def}
% \subsection{Pauli Group Functions}
% \ExecuteMetaData[agda/Data/Pauli.tex]{bad-eq}
\subsection{Fin} \label{agda-fin}
Injectivity of the \(\textbf{Fin}\) type constructor.
\ExecuteMetaData[agda/Data/Fin/Properties.tex]{fin-inj}
\section{References}
\printbibliography[heading=none]
\end{subappendices}
\end{document}

>>>>

\input{preamble}
\title{Finiteness, Cardinality, and Combinatorics in Homotopy Type Theory}
\author{Donnacha Oisín Kidney}
\authorrunning{D. O. Kidney}
\institute{University College Cork \and
  \email{o.kidney@cs.ucc.ie}}
\begin{document}
\maketitle
\begin{abstract}
  We explore five notions of finiteness in Homotopy Type Theory \cite{hottbook}.
  We prove closure properties about all of these notions, culminating in a proof
  that decidable Kuratowski-finite sets form a topos.

@@

  We formalise our work in Cubical Agda
  \cite{vezzosiCubicalAgdaDependently2019}, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
\section{Introduction}
Proofs in a constructive setting are more substantial things than their
classical counterparts.
Rather than just evidence for some fact, they are objects in their own right,
which we can manipulate and extract information from.

@@

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsubsection{Equivalence}
While manifest Bishop finiteness may seem stricter than split enumerability, we
will see soon that it is actually equivalent in strength.
Why study it separately to split enumerability at all, then?
What purpose does it serve on our road to Kuratowski finiteness?
Those questions are answered by the following theorem:
\begin{lemma} \label{bishop-equiv}
  Any manifestly Bishop finite type has an equivalence with a finite prefix of
  the natural numbers.
\end{lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite{hottbook}. \todo[noline]{Need this in appendix also}
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.

@@

Finally, we can prove isomorphisms between the curried and uncurried versions of
functions:
\ExecuteMetaData[agda/Data/Product/N-ary.tex]{curry-n-iso}
And dependent functions:
\ExecuteMetaData[agda/Data/Product/N-ary.tex]{curry-n-pi-iso}
With these combinators, we can implement the search functions in an
arity-generic way:
\ExecuteMetaData[agda/Cardinality/Finite/Search.tex]{all-nary}
And automate away our proofs:
\ExecuteMetaData[agda/Data/Pauli.tex]{prf-curried}
\bibliographystyle{splncs04}
\bibliography{bibliography.bib}
\end{document}
